---
title: "Monte Carlo Integration Examples"
author: "Uma Ravat"
date: "PSTAT 194CS"
output:
  html_document:
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = F}
bfcolor <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{\\textbf{%s}}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'><b>%s</b></span>", color, x)
  } else x
}
```

## Exercise 1: 

If $g(x)=\frac{1}{\sqrt{2\pi}}e^{-x^{2}/2}, \quad -\infty < x < \infty$

**Question 1**: Do you recognize this density function? Which density is it?  What are the parameters? 

`r bfcolor("Solutions:", "blue")`

Write your solution here 

```{r}
# Write any code here
```

**Question 2** Let $\theta$ be the probability that a standard normal variable falls between -1.96 and 1.96. Write this probability as an integral(use latex code to write the integral). Plot the density and shade the area that represents this probability. 


`r bfcolor("Solutions:", "blue")`

Write your solution here 

```{r}
# Write any code here
```

**Question 3** What is the value of $\theta$ from your previous knowledge? 

`r bfcolor("Solutions:", "blue")`

Write your solution here 

```{r}
# Write any code here
```

**Question 4** Approximate this familiar number using Monte Carlo methods in two ways (ie as ratio of number of sampled points and as a mean.)

`r bfcolor("Solutions:", "blue")`
**Method 1:**


Write your solution here 

```{r}
# Write any code here
```


**Method 2:**


Write your solution here 

```{r}
# Write any code here
```


**Question 5** Discuss the difference in the two approaches. Is one better than the other. Why? 

`r bfcolor("Solutions:", "blue")`

Write your solution here 

```{r}
# Write any code here
```

## Exercise 2:

**Question 1:** Consider $\theta = \int_{0}^{1}\frac{1}{2+sin(1/y)}dy$. Compute a confidence interval for $\theta$.

`r bfcolor("Solutions:", "blue")`

Write your solution here 

```{r}
# Write any code here
```

**Question 2:** Come up with two estimators $\theta_1$ and $\theta_2$ for $\theta$. Compare their efficiency. Which one is more efficient and why? 

`r bfcolor("Solutions:", "blue")`
Write your solution here 

```{r}
# Write any code here
```
