---
title: "Monte Carlo Inference Activity"
subtitle: "Exploring Robust Estimators through Simulation"
author: "PSTAT 194CS"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
library(tidyverse)
library(knitr)
library(kableExtra)
# Note: install.packages("moments") if you don't have it
if (!requireNamespace("moments", quietly = TRUE)) {
  cat("Note: moments package not installed. Skewness calculation will be skipped.\n")
}
```

# Introduction

In this activity, you'll conduct your own Monte Carlo studies to understand how different estimators perform under contamination. You'll work through three exercises, building from guided exploration to independent investigation.

## Learning Objectives

By the end of this activity, you will be able to:

1. ✓ Design and implement Monte Carlo simulation studies
2. ✓ Compare estimator performance using MSE
3. ✓ Interpret simulation results in context
4. ✓ Formulate and test statistical hypotheses empirically

## The Contamination Model

Throughout these exercises, we'll work with the contamination model:

$$pN(0, 1) + (1-p)N(0, 100)$$

This represents data where:

- A proportion $p$ of observations come from $N(0,1)$ (good data)
- A proportion $(1-p)$ come from $N(0,100)$ (contaminated/outliers)
- True mean we're trying to estimate: **0**

---

# Exercise 1: Trimmed Mean Estimator

## Part A: Understanding the Code

First, let's review the function for computing MSE of the trimmed mean:

```{r}
# Function to compute MSE of k-level trimmed mean
trimmed_mse <- function(n, m, k, p) {
  # n = sample size per iteration
  # m = number of MC iterations
  # k = trim level (removes k smallest and k largest)
  # p = proportion of good data (1-p is contamination rate)
  
  tmean <- numeric(m)
  
  for(i in 1:m) {
    # Generate contaminated sample
    sigma <- sample(c(1, 10), size = n, replace = TRUE, 
                    prob = c(p, 1-p))
    x <- rnorm(n, 0, sigma)
    x_sorted <- sort(x)
    
    # Compute trimmed mean
    if(k == 0) {
      tmean[i] <- mean(x_sorted)  # Regular mean
    } else {
      tmean[i] <- mean(x_sorted[(k+1):(n-k)])  # Trimmed mean
    }
  }
  
  # MSE calculation (true mean = 0)
  mse <- mean(tmean^2)
  se_mse <- sd(tmean^2) / sqrt(m)
  
  return(c(mse = mse, se = se_mse))
}
```

### 🤔 Understanding Check (discuss with your group):

1. **Why do we use `tmean^2` to compute MSE?**
   
   *Hint: What is MSE when the true parameter is 0?*
   
   <details>
   <summary>Click for answer</summary>
   MSE = E[(θ̂ - θ)²]. When θ = 0, this simplifies to E[θ̂²], which we estimate with mean(tmean^2).
   </details>

2. **What does k=0 represent?**
   
   <details>
   <summary>Click for answer</summary>
   k=0 means no trimming, so we get the regular sample mean.
   </details>

3. **If n=20 and k=5, how many observations are we using?**
   
   <details>
   <summary>Click for answer</summary>
   We remove 5 from each end, leaving 20-10=10 observations for the mean.
   </details>

## Part B: Run the Basic Study

Let's replicate the analysis from class:

```{r basic_study, cache=TRUE}
set.seed(194)  # For reproducibility

# Study parameters
n <- 20        # Sample size
m <- 10000     # MC iterations
K <- n/2 - 1   # Maximum trim level

# Storage for results
results <- expand.grid(k = 0:9, p = c(1, 0.95, 0.9))
results$mse <- NA
results$se <- NA

# Run simulations
for(i in 1:nrow(results)) {
  res <- trimmed_mse(n, m, results$k[i], results$p[i])
  results$mse[i] <- res[1]
  results$se[i] <- res[2]
}

# View results
results %>%
  mutate(contamination = paste0((1-p)*100, "%"),
         mse = round(mse, 4),
         se = round(se, 5)) %>%
  select(k, contamination, mse, se) %>%
  pivot_wider(names_from = contamination, 
              values_from = c(mse, se)) %>%
  kable(caption = "MSE (and SE) by Trim Level and Contamination") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Part C: Visualize Results

```{r plot_results, fig.height=5}
ggplot(results, aes(x = k, y = mse, color = factor(p), group = p)) +
  geom_line(size = 1.5) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mse - 2*se, ymax = mse + 2*se), 
                width = 0.2, alpha = 0.5) +
  scale_color_manual(values = c("green3", "orange", "red"),
                     labels = c("No contamination (p=1)", 
                               "5% contamination (p=0.95)", 
                               "10% contamination (p=0.9)")) +
  labs(title = "MSE of Trimmed Mean Estimators",
       subtitle = "Error bars show ±2 SE",
       x = "Trim Level (k)", 
       y = "Mean Squared Error",
       color = "Scenario") +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))
```

## Part D: Interpret the Results

### Questions for Discussion:

1. **What happens to MSE as we increase k when there's NO contamination (p=1)?**
   
   *Your answer:*
   
   ```
   
   
   
   ```

2. **What trim level (k) gives the best (lowest) MSE for 5% contamination?**
   
   *Your answer:*
   
   ```
   
   
   
   ```

3. **Why does this optimal trim level make sense?**
   
   *Hint: Think about how many contaminated observations you expect in a sample of 20.*
   
   *Your answer:*
   
   ```
   
   
   
   ```

---

# Exercise 2: Sample Median vs Trimmed Mean

## Part A: Formulate Your Hypothesis

Before running any code, think about how the sample median might perform.

### 🧠 Intuition Building:

1. **The median is equivalent to what trim level?**
   
   *Hint: For n=20, the median uses which observations?*
   
   *Your answer:*
   
   ```
   
   
   ```

2. **Prediction: Will the median be more or less robust than low-trim estimators?**
   
   *Your reasoning:*
   
   ```
   
   
   
   ```

## Part B: Implement Median Comparison

Now modify the code to include the median:

```{r median_study, cache=TRUE}
# Function to compute MSE for median
median_mse <- function(n, m, p) {
  med <- numeric(m)
  
  for(i in 1:m) {
    sigma <- sample(c(1, 10), size = n, replace = TRUE, 
                    prob = c(p, 1-p))
    x <- rnorm(n, 0, sigma)
    
    med[i] <- median(x)
  }
  
  mse <- mean(med^2)
  se_mse <- sd(med^2) / sqrt(m)
  
  return(c(mse = mse, se = se_mse))
}

# Add median results to our previous results
set.seed(194)
median_results <- data.frame(
  k = rep("Median", 3),
  p = c(1, 0.95, 0.9)
)

for(i in 1:3) {
  res <- median_mse(20, 10000, median_results$p[i])
  median_results$mse[i] <- res[1]
  median_results$se[i] <- res[2]
}

# Combine with previous results
results_all <- results %>%
  mutate(estimator = paste0("Trim k=", k),
         k = as.character(k)) %>%
  bind_rows(median_results %>% mutate(estimator = "Median")) %>%
  mutate(contamination = (1-p)*100)

# Plot comparison
results_all %>%
  filter(contamination %in% c(0, 5, 10)) %>%
  ggplot(aes(x = reorder(estimator, mse), y = mse, fill = factor(contamination))) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("green3", "orange", "red"),
                    labels = c("0%", "5%", "10%")) +
  labs(title = "Comparing All Estimators",
       x = "Estimator", y = "MSE", fill = "Contamination") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
```

## Part C: Analyze and Discuss

Create a table comparing the median to other estimators:

```{r}
comparison_table <- results_all %>%
  filter(contamination == 5) %>%
  select(estimator, mse, se) %>%
  arrange(mse) %>%
  mutate(mse = round(mse, 4),
         se = round(se, 5),
         rank = row_number())

comparison_table %>%
  kable(caption = "Estimator Rankings for 5% Contamination") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  row_spec(which(comparison_table$estimator == "Median"), 
           bold = TRUE, background = "yellow")
```

### Discussion Questions:

1. **Where does the median rank?**
   
   *Your answer:*
   
   ```
   
   
   ```

2. **Did this match your prediction from Part A?**
   
   *Your answer:*
   
   ```
   
   
   ```

3. **The median is a special case of which trimmed mean?**
   
   *Your answer:*
   
   ```
   
   
   ```

4. **Would the median work well for ANY distribution? Why or why not?**
   
   *Think about: What if the true distribution is very skewed?*
   
   *Your answer:*
   
   ```
   
   
   
   ```

## Part D: Theoretical Verification

For the uncontaminated case, we can compare to theory:

```{r theoretical_comparison}
# Theoretical values for N(0,1) with n=20
n <- 20

# Mean
var_mean <- 1/n  # σ²/n
mse_mean_theory <- var_mean

# Median (approximate for normal)
var_median_approx <- (pi/2) * (1/n)  # ≈ 1.57/n
mse_median_theory <- var_median_approx

cat("Theoretical MSE (no contamination, n=20):\n")
cat("Mean:  ", round(mse_mean_theory, 4), "\n")
cat("Median:", round(mse_median_theory, 4), "\n\n")

cat("Our MC Estimates:\n")
cat("Mean (k=0):", round(results_all$mse[results_all$estimator == "Trim k=0" & 
                                          results_all$contamination == 0], 4), "\n")
cat("Median:    ", round(results_all$mse[results_all$estimator == "Median" & 
                                          results_all$contamination == 0], 4), "\n")
```

**Do they match? If not, why might there be small differences?**

---

# Exercise 3: Effect of Increased Contamination

## Part A: Make Predictions

You're now going to study what happens with **40% contamination** (p=0.6).

### 🎯 Before coding, predict:

1. **Will trimming help more or less than with 5% contamination?**
   
   *Your hypothesis:*
   
   ```
   
   
   ```

2. **What trim level do you think will be optimal?**
   
   *Your reasoning:*
   
   ```
   
   
   ```

3. **How will the regular mean (k=0) perform?**
   
   *Your prediction:*
   
   ```
   
   
   ```

## Part B: Run the Study

```{r high_contamination, cache=TRUE}
set.seed(194)

# YOUR CODE HERE
# Modify the parameters to study p=0.6 (40% contamination)
# Compare to previous results

# Hint: You can copy and modify the code from Exercise 1

# Study parameters
n <- 20
m <- 10000

# Run for p=0.6
high_contam_results <- expand.grid(k = 0:9, p = 0.6)
high_contam_results$mse <- NA
high_contam_results$se <- NA

for(i in 1:nrow(high_contam_results)) {
  res <- trimmed_mse(n, m, high_contam_results$k[i], high_contam_results$p[i])
  high_contam_results$mse[i] <- res[1]
  high_contam_results$se[i] <- res[2]
}

# Add median
median_40 <- median_mse(20, 10000, 0.6)
```

## Part C: Visualization and Comparison

```{r compare_contamination_levels, fig.height=6}
# Combine all contamination levels
all_levels <- results %>%
  mutate(k = as.character(k)) %>%
  bind_rows(high_contam_results %>% mutate(k = as.character(k))) %>%
  mutate(contamination_pct = (1-p)*100)

# Plot all levels
ggplot(all_levels, aes(x = as.numeric(k), y = mse, 
                       color = factor(contamination_pct), 
                       group = contamination_pct)) +
  geom_line(size = 1.5) +
  geom_point(size = 3) +
  scale_color_manual(values = c("green3", "orange", "red", "purple"),
                     name = "Contamination") +
  labs(title = "Impact of Contamination Level on Optimal Trimming",
       x = "Trim Level (k)", 
       y = "Mean Squared Error") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Part D: Analysis Questions

1. **What trim level is optimal for 40% contamination?**
   
   ```{r}
   # Find optimal k
   optimal_k <- high_contam_results$k[which.min(high_contam_results$mse)]
   cat("Optimal k for 40% contamination:", optimal_k, "\n")
   cat("MSE at optimal k:", 
       round(min(high_contam_results$mse), 4), "\n")
   ```
   
   *Your interpretation:*
   
   ```
   
   
   ```

2. **How much worse is the regular mean (k=0) at 40% vs 5% contamination?**
   
   ```{r}
   # YOUR CODE: Compare MSE of k=0 at different contamination levels
   
   
   ```

3. **Create a summary table:**
   
   ```{r}
   # Table showing optimal trim level for each contamination rate
   summary_optimal <- all_levels %>%
     group_by(contamination_pct) %>%
     summarize(
       optimal_k = k[which.min(mse)],
       min_mse = min(mse),
       mean_k0_mse = mse[k == "0"]
     ) %>%
     mutate(benefit_of_trimming = mean_k0_mse / min_mse)
   
   summary_optimal %>%
     mutate(across(where(is.numeric), ~round(., 3))) %>%
     kable(caption = "Optimal Trimming Strategy by Contamination Level") %>%
     kable_styling(bootstrap_options = c("striped", "hover"))
   ```

4. **Did your predictions from Part A match the results?**
   
   *Reflection:*
   
   ```
   
   
   
   ```

---

# Exercise 4: Design Your Own Study! 🚀

Now it's time to explore something that interests you!

## Part A: Choose Your Research Question

Here are some ideas (or create your own):

### Option 1: Different Distributions
What if the uncontaminated data isn't normal? Try:
- Exponential distribution
- Uniform distribution
- t-distribution with low df

### Option 2: Different Estimators
Compare other robust estimators:
- Winsorized mean (replace extremes instead of removing)
- Interquartile mean (trim to 25th-75th percentiles)
- Huber M-estimator

### Option 3: Different Contamination Patterns
What if contamination isn't symmetric?
- One-sided contamination (only large values)
- Multiple contamination sources
- Increasing contamination with sample size

### Option 4: Practical Application
Apply to real data:
- Stock returns with crashes
- Sensor data with malfunctions
- Survey data with trolls

**Your chosen question:**

```



```

## Part B: Design Your Study

### Study Design Plan:

1. **What are you testing?**
   
   ```
   
   
   ```

2. **What parameters will you vary?**
   
   ```
   
   
   ```

3. **How will you measure performance?**
   
   ```
   
   
   ```

4. **What do you expect to find?**
   
   ```
   
   
   ```

## Part C: Implement Your Study

```{r your_study, cache=TRUE}
# YOUR CODE HERE
# Design and implement your Monte Carlo study

set.seed(194)  # Keep for reproducibility

# Example template:
custom_study <- function() {
  # Define your parameters
  
  # Run simulations
  
  # Calculate and return results
  
}

# Run your study

# Create visualizations

```

## Part D: Results and Interpretation

### Findings:

```{r}
# Create tables and plots summarizing your results


```

### Interpretation:

**What did you learn?**

```



```

**Did the results surprise you?**

```



```

**What are the practical implications?**

```



```

### Theory Connection:

**Are there theoretical results that support/explain your findings?**

```



```

---

# Reflection and Synthesis

## Group Discussion Questions:

1. **Across all exercises, what's the key trade-off with trimmed means?**
   
   *Your group's answer:*
   
   ```
   
   
   ```

2. **In practice, how would you choose a trim level without knowing the true contamination rate?**
   
   *Ideas:*
   
   ```
   
   
   ```

3. **What are the limitations of the contamination model we used?**
   
   *Your thoughts:*
   
   ```
   
   
   ```

4. **How could you extend this methodology to real data analysis?**
   
   *Applications:*
   
   ```
   
   
   ```

---

# Bonus Challenges

If you finish early, try these extensions:

## Challenge 1: Confidence Intervals

Compute confidence intervals for the MSE estimates and see if theoretical values fall within them.

```{r eval=FALSE}
# Template
compute_mse_ci <- function(estimates, true_value, conf_level = 0.95) {
  # YOUR CODE
}
```

## Challenge 2: Power Analysis

How many MC iterations (m) do you need to reliably detect a 10% difference in MSE between two estimators?

## Challenge 3: Bias-Variance Decomposition

Decompose MSE into bias² and variance for each estimator. Which dominates?

```{r eval=FALSE}
# Template
bias_variance_decomp <- function(estimates, true_value) {
  bias <- mean(estimates) - true_value
  variance <- var(estimates)
  
  list(
    bias_squared = bias^2,
    variance = variance,
    mse = bias^2 + variance
  )
}
```

---

# Summary Presentation (3 minutes)

Prepare a brief presentation for the class:

## Slide 1: Research Question
- What did you investigate?
- Why is it interesting?

## Slide 2: Methods
- How did you design your MC study?
- What parameters did you test?

## Slide 3: Key Results
- Show your main plot/table
- Highlight most interesting finding

## Slide 4: Conclusions
- What did you learn?
- Practical implications?
- Connection to theory?

---

# Additional Resources

## Further Reading

1. **Robust Statistics:**
   - Huber, P. J. (1981). *Robust Statistics*
   - Hampel, F. R., et al. (1986). *Robust Statistics: The Approach Based on Influence Functions*

2. **Monte Carlo Methods:**
   - Rizzo, M. L. (2019). *Statistical Computing with R* - Chapter 6
   - Robert, C. P., & Casella, G. (2004). *Monte Carlo Statistical Methods*

3. **Real Applications:**
   - Deliveroo article: https://deliveroo.engineering/2018/12/07/monte-carlo-power-analysis.html

## R Packages to Explore

```{r eval=FALSE}
# Robust statistics
install.packages("robustbase")
install.packages("MASS")

# Visualization
install.packages("ggplot2")
install.packages("plotly")  # Interactive plots

# Power analysis
install.packages("pwr")
```

---

# Submission Checklist

Before you finish, make sure you've:

- [ ] Completed all three main exercises
- [ ] Designed and run your own study (Exercise 4)
- [ ] Interpreted all results (not just reported numbers)
- [ ] Connected findings to theoretical concepts
- [ ] Created clear visualizations
- [ ] Prepared your 3-minute presentation
- [ ] Discussed as a group and recorded insights

**Turn in:** This completed worksheet + presentation slides

---

*Good luck and have fun exploring! Remember: The goal isn't just to get the right answer, but to understand the process of using Monte Carlo methods to answer statistical questions.*
